---
title: "Airline_Delay_Cause"
authors: "Annie Chan, Halle Davis, Salvador R. Sanchez Castro"
date: '2022-11-07'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Arriving 2 Hours Early, Staying 3 Hours Late - Time-Series Analysis of Flight Delay**

**Team Number**: 6

**Team Leader/Representative**: Halle Davis

**Full Names of Team Members**: 1. Sanchez Castro Salvador 2. Lai Ieng Chan 3. Halle Davis

------------------------------------------------------------------------

## Libraries and Data

```{r Libs}
library(visdat)
library(DataExplorer)
library(dplyr)
library(zoo)
library(ggplot2)
library(plotly)
library(htmlwidgets)
library(tidyverse)
library(fpp2)
library(readr)
library(moments)
library(caret)
library(outliers)
library(TSstudio)

set.seed(506)
```

```{r Load Data}
Airline_Delays <- read.csv('Airline_Delay_Cause.csv')
```

## EDA

### Initial Peek into Data

```{r peek}
head(Airline_Delays)
```

```{r EDA}
glimpse(Airline_Delays)
cat("\n")
cat("\n")
summary(Airline_Delays)
```

```{r Counts}
cat("Dimensions: ", dim(Airline_Delays))
cat("\n")
cat("Number of Carriers: ", length(unique(Airline_Delays$carrier)))
cat("\n")
cat("Number of Airports: ", length(unique(Airline_Delays$airport)))
```

```{r DataExplorer}
#Hash out to improve when running
#DataExplorer::create_report(Airline_Delays)
```

### Categorical Variables Visualized

```{r Categorical Variables}

## Carrier
barplot(table(Airline_Delays$carrier), main="Bar Graph of Carrier",
        ylab="Frequency", xlab = "Carrier")
#Didn't visualize carrier name because its redundant to carrier

## Airport
barplot(table(Airline_Delays$airport), main="Bar Graph of Airport",
        ylab="Frequency", xlab = "Airport")
#Didn't visualize airport name because its redundant to airport

## Year
barplot(table(Airline_Delays$year), main="Bar Graph of Year",
        ylab="Frequency", xlab = "Year")

## Month
barplot(table(Airline_Delays$month), main="Bar Graph of Month",
        ylab="Frequency", xlab = "Month")


```

### Numerical Variables Visualized

```{r Numerical Variables}

## arr_flights
hist(Airline_Delays$arr_flights, main="Histogram of Arriving Flights",
     ylab="Frequency", xlab = "Arriving Flights")

## arr_del15
hist(Airline_Delays$arr_del15, main="Histogram of Delayed Flights",
     ylab="Frequency", xlab = "Delayed Flights")

## arr_cancelled
hist(Airline_Delays$arr_cancelled, main="Histogram of Cancelled Flights",
     ylab="Frequency", xlab = "Cancelled Flights")

## arr_diverted
hist(Airline_Delays$arr_diverted, main="Histogram of Diverted Flights",
     ylab="Frequency", xlab = "Diverted Flights")

## carrier_ct
hist(Airline_Delays$carrier_ct, main="Histogram of Flights Delayed due to Carrier",
     ylab="Frequency", xlab = "Carrier Delays")

## carrier_delay
hist(Airline_Delays$carrier_delay, main="Histogram of Time Delayed due to Carrier",
     ylab="Frequency", xlab = "Carrier Delay Time")

## weather_ct
hist(Airline_Delays$weather_ct, main="Histogram of Flights Delayed due to Weather",
     ylab="Frequency", xlab = "Weather Delays")

## weather_delay
hist(Airline_Delays$weather_delay, main="Histogram of Time Delayed due to Weather",
     ylab="Frequency", xlab = "Weather Delay Time")

## nas_ct
hist(Airline_Delays$nas_ct, main="Histogram of Flights Delayed due to NAS",
     ylab="Frequency", xlab = "NAS Delays")

## nas_delay
hist(Airline_Delays$nas_delay, main="Histogram of Time Delayed due to NAS",
     ylab="Frequency", xlab = "NAS Delay Time")

## security_ct
hist(Airline_Delays$security_ct, main="Histogram of Flights Delayed due to Security",
     ylab="Frequency", xlab = "Security Delays")

## security_delay
hist(Airline_Delays$security_delay, main="Histogram of Time Delayed due to Security",
     ylab="Frequency", xlab = "Security Delay Time")

## late_aircraft_ct
hist(Airline_Delays$late_aircraft_ct, main="Histogram of Flights Delayed due to Late Aircraft",
     ylab="Frequency", xlab = "Late Aircraft Delays")

## late_aircraft_delay
hist(Airline_Delays$late_aircraft_delay, main="Histogram of Time Delayed due to Late Aircraft",
     ylab="Frequency", xlab = "Late Aircraft Delay Time")

```

## Pre-Processing

### Date Transformation

```{r Date Column}
#To year-Date
Airline_Delays$date <- as.yearmon(paste(Airline_Delays$year,
                                        Airline_Delays$month), "%Y %m")
#To Date type
Airline_Delays <- transform(Airline_Delays, date = as.Date(date))
Airline_Delays 
```

### Filtering to One Airport (Based on # of Observations)

```{r Selecting Airport and Airline}

airport_table <- table(Airline_Delays$airport)
airport_table <- airport_table[order(airport_table,decreasing = TRUE)]

#Airports with most flights
airport_table

```
### Counts by Air Carrier

```{r Filter for Airport}

Airline_Delays_Pit <- filter(Airline_Delays , airport == "PIT")
Airline_Delays_Pit


Airline_Delays_DFW <- filter(Airline_Delays , airport == "DFW")
Airline_Delays_DFW

carrier_table <- table(Airline_Delays_Pit$carrier)
carrier_table <- carrier_table[order(carrier_table,decreasing = TRUE)]

#Airports with most flights
carrier_table

sum <- summary(Airline_Delays_Pit$date)
sum
```

### Checking for Duplicates

```{r}

sum(duplicated(Airline_Delays_Pit))

```

### Checking for Nulls

```{r Checking for Nulls}

sum(sapply(Airline_Delays_Pit, function(x) sum(is.na(x))))

```

### Checking for Skew

```{r Checking for Skew}

nums <- unlist(lapply(Airline_Delays_Pit, is.numeric), use.names = FALSE)  
Airline_Delays_Pit_Nums <- Airline_Delays_Pit[ , nums]

sapply(Airline_Delays_Pit_Nums, function(x) skewness(x))

```

### Checking for Near Zero Variance

```{r Checking for NZV}

length(nearZeroVar(Airline_Delays_Pit))
degeneratecols <- nearZeroVar(Airline_Delays_Pit)
degeneratecols #Airport, Airport Name, Security Counts, Security Delays

```

### Checking for MultiCollinearity

```{r Checking for MultiCollinearity}

highCorr <- findCorrelation(cor(Airline_Delays_Pit_Nums), cutoff = .75)
highCorr

```

### Checking for Outliers

```{r Checking for Outliers}

boxplot(Airline_Delays_Pit$arr_delay,
        main = "Boxplot for Delays",
        xlab = 'PIT',
        ylab = 'Airport Delay in Minutes'
        )

subset(Airline_Delays_Pit, arr_flights < arr_del15)

```
While it seems that there are outliers based on the simple 1.5 * the IQR rule
for the number of delayed flights (our dependent variable of interest), 
there is no case where the number of delayed flights is greater than the number
of arriving flights, meaning that all the data is reasonable and likely to be
meaningful.


### Visualizing Time Series -- Delays over Time

```{r view time series}
arr_del15_pit.ts <- ts(Airline_Delays_Pit$arr_del15, ## arr_del15 is the number of flights that were delayed by                                                        at least 15 minutes
                 start = c(2019, 1),
                 end = c(2022, 8),
                 freq = 12)

#Sum Delays by month
Airline_Delays_Pit_sum <- Airline_Delays_Pit %>%
  group_by(date) %>%
  summarise(Freq = sum(arr_delay))

Airline_Delays_Dfw_sum <- Airline_Delays_DFW %>%
  group_by(date) %>%
  summarise(Freq = sum(arr_delay))

## arr_delay is the total amount of delay in minutes
arr_delays_pit.ts <- ts(Airline_Delays_Pit_sum$Freq  , 
                 start = c(2019, 1),
                 end = c(2022, 8),
                 freq = 12)

arr_delays_dfw.ts <- ts(Airline_Delays_Dfw_sum$Freq  , ## arr_delay is the total amount of delay in minutes
                 start = c(2019, 1),
                 end = c(2022, 8),
                 freq = 12)


autoplot(arr_delays_pit.ts,
         series = 'PIT Delays',
         xlab = 'Date',
         ylab = 'Airport Delay in Minutes',
         main = 'Delays for Pittsburgh International')

autoplot(arr_delays_dfw.ts,
         series = 'DFW Delays',
         xlab = 'Date',
         ylab = 'Airport Delay in Minutes',
         main = 'Delays for Dallas/Fort Worth International')
```

### Visualizing Time Series -- Broken Down by Carrier

```{r view time series}
fig <- plot_ly(Airline_Delays_Pit,
               type = 'scatter',
               mode = 'lines'
               #line = list(color = '#326ac2')
               )%>%
  add_trace(x = ~date,
            y = ~arr_del15, ## changed to airdel15
            color = ~carrier_name)%>%
  layout(showlegend = T)
  options(warn = -1)

fig <- fig %>%   layout(
         xaxis = list(
           title=list(text='Time',
                      font = list(size = 20),
                      standoff = 25),
           tickangle=-45
           ),
         yaxis = list(
           title=list(text='Airlport Delays',
                      font = list(size = 20),
                      standoff = 25),
           plot_bgcolor='#e5ecf6', width = 900)
           )
fig


saveWidget(fig, "Airline_Delays_Pit.html", selfcontained = F, libdir = "lib")

```


### Finding Relationship between Dependent and Independent

```{r Statistical Tests of Significance}

## Month
chisq.test(Airline_Delays_Pit$month, Airline_Delays_Pit$arr_del15,
           correct=FALSE)
#Not a significant relationship between month and delays

## Year
chisq.test(Airline_Delays_Pit$year, Airline_Delays_Pit$arr_del15,
           correct=FALSE)

t.test(Airline_Delays_Pit$year,
       Airline_Delays_Pit$arr_del15)
#Significant relationship between year and delays

```

```{r Autocorrelation Plots}

arr_delays_pit.ts  %>% 
  acf(lag.max = 40, 
      main = "Autocorrelation Plot - R")

# pacf R time series
arr_delays_pit.ts %>%
  pacf(lag.max = 40,
       main = "Partial Autocorrelation Plot - R")

```

### Partitioning Data to Training and Validation

```{r Spliting Data}

Airline_Delays_Pit.train <- window(arr_delays_pit.ts, end = c(2022,3))
Airline_Delays_Pit.test <- window(arr_delays_pit.ts, start = c(2022,4))

autoplot(Airline_Delays_Pit.train, series = 'Train',
         xlab = 'Date',
         ylab = 'Airport Delay in Minutes',
         main = 'Train/Test Time Series Split')  +
  autolayer(Airline_Delays_Pit.test, series = 'Test')
```
## Modeling

```{r Modeling Variables}

forecast.periods <- 5
y <- Airline_Delays_Pit.train
test.y <- Airline_Delays_Pit.test

```

```{r naive}
#Naive
naive.model <- naive(y, h = forecast.periods)

#SeasonalNaive
naive.season.model <- snaive(y, h = forecast.periods)

#Mean
mean.model <- meanf(y, h = forecast.periods)

#Plot
autoplot(y, series = 'Training',
         xlab = 'Date',
         ylab = 'Airport Delay in Minutes',
         main = 'Naive Model') +
        autolayer(Airline_Delays_Pit.test, series = 'Validation') +
        autolayer(naive.model, alpha = .3, series = 'Naive') 

accuracy(naive.model, Airline_Delays_Pit.test)

autoplot(y, series = 'Training') +
  autolayer(Airline_Delays_Pit.test, series = 'Validation') +
  autolayer(naive.season.model, alpha = .3, series = 'Seasonal Naive') 

accuracy(naive.season.model, Airline_Delays_Pit.test)

autoplot(y, series = 'Training') +
  autolayer(Airline_Delays_Pit.test, series = 'Validation') +
  autolayer(mean.model, alpha = .3, series = 'Mean') 

accuracy(mean.model, Airline_Delays_Pit.test)

```



```{r ETS}

# Simple Exponential Model
ets.model <- ets(y,
                 model = "ANN",
                 alpha = 0.9,
                 gamma = .05,
                 restrict = FALSE
                 )
ets.model

ets.pred <- forecast(ets.model, h=forecast.periods)

accuracy(ets.pred, Airline_Delays_Pit.test)

#Plot
autoplot(y, series = 'Training') +
  autolayer(Airline_Delays_Pit.test, series = 'Validation') +
  autolayer(ets.pred, alpha = .3, series = 'Forecast') 

```



```{r auto arima}

#Autoarima
auto_fit <- auto.arima(y)
summary(auto_fit)

#Residuals
checkresiduals(auto_fit)

#Forecast
auto.fore <- forecast(auto_fit, h=forecast.periods)
fore <- forecast(fit, h = forecast.periods)

#Plot
autoplot(y, series = 'Training') +
  autolayer(Airline_Delays_Pit.test, series = 'Validation') +
  autolayer(auto.fore, alpha = .3, series = 'Auto Forecast')

#autoplot(auto_fit)
accuracy(auto.fore, Airline_Delays_Pit.test)
```

```{r arima tunning}
acf(y, lag.max = 300)
acf(diff(y, lag = 12), lag.max = 72)
acf(diff(diff(y, lag = 12)), lag.max = 72)

pacf(y, lag.max = 72)
pacf(diff(y, lag = 12), lag.max = 72)
pacf(diff(diff(y, lag = 12)), lag.max = 72)

autoplot(diff(diff(y, lag = 12)))
```



```{r arima}

#Arima
fit <- arima(y , order=c(0,1,1), seasonal = c(0,1,1))

summary(fit)

#Residuals
checkresiduals(fit)

#Forecast
fore <- forecast(fit, h = forecast.periods)

#Plot
autoplot(y, series = 'Training',
         xlab = 'Date',
         ylab = 'Airport Delay in Minutes',
         main = 'Arima Model (0,1,1), (0,1,1)') +
  autolayer(Airline_Delays_Pit.test, series = 'Validation')  +
  autolayer(fore, alpha = .3, series = 'Forecast') 

#autoplot(fit)
accuracy(fore, Airline_Delays_Pit.test)
```
```{r covid Dummy}
y
y.frame <- data.frame(Y=as.matrix(y),date=as.Date(as.yearmon(time(y))))
y.frame.covid <- y.frame  %>%
  mutate(Covid = ifelse(date > '2020-02-01' & date < '2021-06-01' , 1, 0))
y.frame.covid

test.frame <- data.frame(Y=as.matrix(test.y),date=as.Date(as.yearmon(time(test.y))))
test.y.frame.covid <- test.frame %>%
  mutate(Covid = ifelse(date > '2020-02-01' & date < '2021-06-01' , 1, 0))
test.y.frame.covid
```

```{r improved arima}

predictors <- as.matrix(y.frame.covid[c("Covid")] )
predictors.valid <- as.matrix(test.y.frame.covid[c("Covid")] )

fit.predictor <- Arima(y,
                            xreg = predictors,
                            order=c(0,1,0),
                            seasonal = c(1,1,1))

forecast <- forecast(fit.predictor,
                     xreg = predictors.valid,
                     h=forecast.periods)

summary(fit.predictor)

autoplot(y, series = 'Training',
         xlab = 'Date',
         ylab = 'Airport Delay in Minutes',
         main = 'Arima Model (0,1,1),(0,1,1) with COVID Predictor')  +
  autolayer(Airline_Delays_Pit.test, series = 'Validation')  +
  autolayer(forecast, alpha = .3, series = 'Forecast')  

accuracy(forecast, Airline_Delays_Pit.test)

#autoplot(fit.predictor)
```



```{r regression}

plane_lm <- tslm(Airline_Delays_Pit.train ~ season)
summary(plane_lm)

plane_lm_predict <- forecast(plane_lm, h = forecast.periods)

accuracy(plane_lm_predict, Airline_Delays_Pit.test)


autoplot(Airline_Delays_Pit.train, series = 'train') +
  autolayer(Airline_Delays_Pit.test, series = 'actual') +
  autolayer(plane_lm_predict, series = 'prediction')

```

```{r neuralnetwork}

plane_nnet <- nnetar(Airline_Delays_Pit.train)
plane_nnet_predict <- forecast(plane_lm, h = forecast.periods)

accuracy(plane_nnet_predict, Airline_Delays_Pit.test)

autoplot(Airline_Delays_Pit.train, series = 'train') +
  autolayer(Airline_Delays_Pit.test, series = 'actual') +
  autolayer(plane_nnet_predict, series = 'prediction')

autoplot(y, series = 'Training',
         xlab = 'Date',
         ylab = 'Airport Delay in Minutes',
         main = 'Neuralnetwork Model')  +
  autolayer(Airline_Delays_Pit.test, series = 'Validation')  +
  autolayer(plane_nnet_predict, alpha = .3, series = 'Forecast')  

```